### 2020-07-16 manuscript/manuscript.rmd
make sure e.g. and et al. are used consistently throughout the manuscript

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. Or for data-partitioning analyses, some gene partitions use different site-homogeneous models. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). Gene sampling varies in several ways, including in the fractions of single copy orthologs (e.g., BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models. In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes. Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.

@@

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson-CAT model.

@@

These results further clarify when analyses support Porifera-sister (e.g., Philippe2009, Whelan2017). Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a simplistic Poisson rather than GTR exchange matrix must be used. Analyses under other conditions either support Ctenophora-sister or are unresolved.

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. Instead of continuing to use previous matrices that has limited taxa sampling (e.g., Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference. [XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling. Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix. We also used a table where manual entries were needed (manual_taxonomy_map.tsv), e.g. authors of the original matrix indicate species name in original manuscript. For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.

@@

#### Phylogenetic analyses in IQ-TREE
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

**Extended Data Fig. 5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

@@

### S.1 Models of molecular evolution
Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).  Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used.

@@

#### Hejnol et al. 2009

Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms. The orthology inference was largely similar to Dunn et al. 2008, with the exception of orthology genes were clustered by MCL. The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy. With the exception of 53 gene matrix, maximum likelihood analyses from all other datasets strongly supported Ctenophora-sister (models were selected by RaxML perl script).

@@

#### Whelan et al. 2015
Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g. long-branch attraction, compositional bias, fast evolving genes, etc.). Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.  The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups. All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister. GTR-CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.

@@

Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity. The authors constructed a dataset with 200 protein markers based on Philippe et al. 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa. Both site-heterogeneous Poisson+CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.

@@

Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets). It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods based on recovering Porifera-sister hypothesis. The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets. The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.

@@

There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig. 5). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 6).

@@

This conclusion does not follow from their own presented results. Only a single analysis with posterior predictive scores provides what could be considered strong support > 95 posterior probability) for Porifera-sister. Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices). Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses (Extended Data Fig. 5). Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister. Of the six recoded analyses, five are unresolved. Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix. Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig. 5). The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig. 5), indicating that it did not converge as well as other analyses.

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data. we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 3B).

>>>>

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. Or for data-partitioning analyses, some gene partitions use different site-homogeneous models. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models. In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes. Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.

@@

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+CAT model.

@@

These results further clarify when analyses support Porifera-sister (*e.g.* Philippe2009, Whelan2017). Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a simplistic Poisson rather than GTR exchange matrix must be used. Analyses under other conditions either support Ctenophora-sister or are unresolved.

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. Instead of continuing to use previous matrices that has limited taxa sampling (*e.g.* Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference. [XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling. Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

We programmatically queried the NCBI Taxonomy database to standardize names of samples in each matrix. We also used a table where manual entries were needed (manual_taxonomy_map.tsv), *e.g.* authors of the original matrix indicate species name in original manuscript. For a table summarizing all samples and their new or lengthened names, see taxon_table.tsv.

@@

#### Phylogenetic analyses in IQ-TREE
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG, LG, GTR models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

**Extended Data Fig. 5.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

@@

### S.1 Models of molecular evolution
Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT).  Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used.

@@

#### Hejnol *et al.* 2009

Hejnol *et al.* [@hejnol2009assessing] added EST sequences from seven taxa, and a total of 94 taxa were included in the final data matrix to explore animal phylogeny, especially the position of acoelomorph flatwroms. The orthology inference was largely similar to Dunn *et al.* 2008, with the exception of orthology genes were clustered by MCL. The final data matrix included 1497 genes, and then subsampled with 844, 330 and 53 gens by different thresholds of gene occupancy. With the exception of 53 gene matrix, maximum likelihood analyses from all other datasets strongly supported Ctenophora-sister (models were selected by RaxML perl script).

@@

#### Whelan et al. 2015
Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (*e.g.* long-branch attraction, compositional bias, fast evolving genes, etc.). Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.  The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups. All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister. GTR+CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.

@@

Chang *et al.* [@Chang:2015hl] was originally used to explore phylogenetic position of Myxozoa in Cnidaria but also sampled broadly across the breadth of animal diversity. The authors constructed a dataset with 200 protein markers based on Philippe *et al.* 2011 [@philippe2011resolving] with 51,940 amino acids and 77 taxa. Both site-heterogeneous Poisson+CAT and site-homogeneous GTR models strongly supported Ctenophora-sister.

@@

Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets). It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods based on recovering Porifera-sister hypothesis. The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (GTR+CAT) models in non-recoding and recoding datasets. The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with GTR+CAT models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in choanoazoa dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.

@@

There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig. 5). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT+G model with PhyloBayes-MPI [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 6).

@@

This conclusion does not follow from their own presented results. Only a single analysis with posterior predictive scores provides what could be considered strong support > 95 posterior probability) for Porifera-sister. Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices). Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister). This leaves eight GTR+CAT analyses (Extended Data Fig. 5). Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister. Of the six recoded analyses, five are unresolved. Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the GTR+CAT analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix. Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Extended Data Fig. 5). The only statistic that stands out for this one analysis is that it has the highest maxdiff (Extended Data Fig. 5), indicating that it did not converge as well as other analyses.

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the GTR+CAT analysis of the SR-6 recoded Whelan data. we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the GTR+CAT model with PhyloBayes-MPI, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 3B).
